## Comparison: `r contrast.name`

```{r dge_intro, results="asis"}
contrast <- params$contrasts[[contrast.name]]
if (length(contrast) == 3) {
  cat("This analysis compares samples where \"", contrast[1], "\" is \"**", contrast[2],
      "**\" to samples where \"", contrast[1], "\" is \"**", contrast[3], "**\" (reference).\n\n", sep="")
  groups <- contrast[-1]
} else if (length(contrast) == 2) {
  cat("This analysis compares condition \"**", contrast[[1]], "**\" to condition \"**", contrast[[2]],
      "**\" (reference).\n\n", sep="")
  groups <- unlist(contrast)
} else if (length(contrast) == 1) {
  cat("This analysis compares condition \"**", contrast[[1]], "**\" to the reference condition.\n\n", sep="")
  groups <- c(contrast[[1]], "reference")
}
cat("- A positive log fold change (\"upregulation\") means: higher gene expression in the \"", groups[1],
    "\" group compared to the \"", groups[2], "\" group.\n\n", sep="")
cat("- A negative log fold change (\"downregulation\") means: higher gene expression in the \"", groups[2],
    "\" group compared to the \"", groups[1], "\" group.\n\n", sep="")
```

```{r dge_res}
res <- results(data.ds, contrast=contrast)
res.ord <- as.data.frame(res[order(res$pvalue), ])
```

### Overview (MA plot)

Points in blue correspond to significant genes (FDR-adjusted p-value < 0.1).

```{r ma_plot}
plotMA(res, cex=1, ylim=c(-2, 2))
```

### Volcano plot

The cut-offs shown in the plot are:

- FDR-adjusted p-value: 0.1

- fold change: `r params$volcano_fc` (up)/`r format(1/params$volcano_fc, digits=2)` (down)

Low-abundance genes without adjusted p-values (due to independent filtering performed by DESeq2) are not included in the plot.

```{r volcano}
## TODO: remove control probes?
ind <- !is.na(res$padj)
EnhancedVolcano(res[ind, ], lab=rownames(res)[ind], x="log2FoldChange", y="padj",
                xlim=range(res$log2FoldChange), ylim=c(0, max(-log10(res[ind, "padj"]), na.rm=TRUE)),
                pCutoff=0.1, FCcutoff=log2(params$volcano_fc), title="", subtitle="",
                xlab=bquote(~log[2] ~ "fold change"), ylab=bquote(~-log[10] ~ p[adj]),
                legendPosition="none", labSize=4)
```

```{r top_genes, eval=(length(contrast) == 3), results="asis", fig.dim=c(10, 7)}
cat("\n\n### Top genes\n\n")
res.part <- res.ord[1:params$plot_top_degs, ]
genes <- rownames(res.part)
samples <- rownames(coldata)[coldata[[contrast[1]]] %in% contrast[-1]]
plot.data <- as.data.frame(norm.counts[genes, samples])
plot.data$gene <- factor(rownames(plot.data), levels=rownames(plot.data)) # preserve order of genes
plot.data <- pivot_longer(plot.data, !gene, names_to="sample", values_to="norm_counts")
plot.data$group <- factor(coldata[plot.data$sample, contrast[1]], levels=c(contrast[3], contrast[2]))

## labels <- paste0(seq_along(genes), ". ", genes, "\nLFC: ", format(res.part$log2FoldChange, digits=2),
##                  ", q: ", format(res.part$padj, digits=2))
## TODO: is this worth it just to get the gene names in bold?
labels <- paste0("atop(\"", seq_along(genes), ".\"~bold(\"", genes, "\"), \"LFC:\"~",
                 format(res.part$log2FoldChange, digits=2), "*\", q:\"~", format(res.part$padj, digits=2), ")")
names(labels) <- genes
plot.data$label <- factor(labels[plot.data$gene], levels=labels) # preserve order

ggplot(plot.data, aes(group, norm_counts + 1)) +
  geom_boxplot() +
  geom_point(alpha=0.5) +
  ## facet_wrap(vars(gene), scales="free_y", labeller=labeller(gene=labels)) +
  facet_wrap(vars(label), scales="free_y", labeller=label_parsed) +
  ## expand_limits(y=1) +
  scale_y_log10() +
  labs(x=contrast[1], y="normalized counts (+1, log-scale)")
```

```{r dge_table}
opts <- list(columnDefs = list(list(orderable=TRUE, targets=0)), # allow sorting by row name (gene)
             order = list(list(5, "asc"))) # order by p-value by default
## TODO: generate different filenames for different contrasts
## show up to 2000 genes (other table becomes unwieldy):
make.table(res.ord[1:min(nrow(res.ord), 2000), ], "dge_results", options=opts, height="100%", width="100%") %>%
  formatRound("baseMean", 1) %>%
  formatRound(2:ncol(res.ord), 3)
```

```{r no_gsea, eval=params$no_gsea}
knitr::knit_exit(fully=FALSE) # stop here
```

### Gene set enrichments {.tabset}

Using a significance threshold of `r params$gsea_pmax` (FDR-adjusted p-value).
Plots show up to 15 top hits.
`(+)` and `(-)` after a term indicate enrichment at the top and bottom, respectively, of the gene list ranked by differential expression.
Typically this corresponds to enrichment among upregulated/downregulated genes (as long as enough changes in both directions are detected -- see gene-level results above).

```{r gsea, results="asis", warning=FALSE}
gsea.file <- file.path(params$out_dir, paste0("GSEA_", paste(unlist(contrast), collapse="-"), ".rds"))
if (params$reuse_output && file.exists(gsea.file)) { # load previous GSEA results
  gsea.res <- readRDS(gsea.file)
} else { # run GSEA
  scores <- get.gsea.input(res)
  gsea.res <- suppressMessages(gsea.all(scores, "", plot=FALSE, reactome.mapping=reactome.mapping,
                                        pvalue.cutoff=params$gsea_pmax))
  saveRDS(gsea.res, file=gsea.file)
}

for (id in names(gsea.res)) {
  cat("\n\n####", gsea.names[id], "\n\n")
  if (nrow(gsea.res[[id]]) == 0) {
    cat("No hits\n\n")
  } else {
    print(dotplot.direction(gsea.res[[id]], label_format=50, font.size=ifelse(id == "Hallmark", 8, 10))) # + labs(title=follow.ups[i]))

    cat("<details>\n<summary>Show table</summary>\n\n") # hide table by default
    ## workaround to print table inside for-loop:
    df <- as.data.frame(gsea.res[[id]])
    opts <- list(order = list(list(6, "asc")), # order by p-value by default
                 pageLength = 5, # show fewer resuls by default
                 scrollX = TRUE, # horizontal scrolling
                 scrollCollapse = TRUE)
    filename <- paste0("GSEA_", sub(".", "-", id, fixed=TRUE))
    cat(knitr::knit_print(make.table(df, filename, options=opts, height="100%", width="100%", rownames=FALSE) %>%
                          formatRound(4:8, 3)))
    ## %>% formatStyle(colnames(df), "white-space"="nowrap")))
    cat("</details>\n\n")
  }
}
```
